Is the direction of greater Granger causal influence the same as the direction of information flow?

### 2015-10-04 paper.tex
Add a reference for comparison of GC.

\begin{document}
\title{Is the direction of Granger causal influence the same as the direction of information flow?}
\author{Praveen Venkatesh and Pulkit Grover\\Carnegie Mellon University}

@@

\begin{abstract}
	Granger causality is an established statistical measure of the ``causal influence'' that one stochastic process has on another. Along with its more recent generalization -- Directed Information -- Granger Causality has been used extensively in neuroscience, and in complex interconnected systems in general, to infer statistical causal influences. Of late, many works have begun to interpret the direction of causal influence as the direction of ``information flow''. We ask: is the direction of causal influence, as predicted by Granger Causality and Directed Information always the same as the direction of information flow? We test whether these measures correctly predict the direction of information flow by using two simple theoretical experiments, in which the true direction of information flow (the ``ground truth'') is known by design. The experiments are based on a communication system with a feedback channel, upon which a strategy inspired by the work of Schalkwijk and Kailath is employed. We show that in these experiments, the direction of information flow can be opposite to that which is inferred using Granger causality and Directed Information. Thus, while it might be reasonable to infer the direction of statistical causal influence using these techniques, one needs to exercise care in interpreting the direction of causal influence as the direction of information flow.
\end{abstract}

@@

\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
	\centering
	\includegraphics[width=3.2in]{gc-comparison}
	\caption{In order to determine the direction of greater causal influence, the Granger Causality metrics in the forward and reverse directions are compared}
	\label{fig:gc-comparison}
\end{figure}

In order to determine the direction of \emph{greater} causal influence, the Granger Causality metric (ratio of residual variances) from $\{X\}$ to $\{Y\}$ is often \emph{compared} to that from $\{Y\}$ to $\{X\}$. The direction of causal influence is then taken to be the direction with the greater Granger Causality metric (e.g.~\cite{Brovelli2004BetaOscillations,jiao2013universal}). Further, this direction of causal influence is interpreted to be the direction of information flow, which is the interpretation we question in this paper. We also note that many of these works use a spectral version of Granger Causality, that supplies this metric as a function of frequency. It then becomes possible to also determine the brain wave frequency at which these influences occur. However, we restrict our analysis to the simpler non-spectral version of Granger Causality, since it is sufficient for the purpose of our arguments.

@@

Suppose we try to use the Granger causality measure of directed influence to infer the direction of information flow. As discussed in Section~\ref{sec:intro}, one might expect the direction of information flow -- and hence the direction of greater Granger causality -- to be from the transmitter to the receiver.

@@

In the reverse direction, the Directed Information is infinite.
\begin{align*}
	I(0*\widehat{\Theta}^{n-1}\rightarrow X^{n}) &= \sum_{i=0}^{n-1}I(X_{i+1};\widehat{\Theta}^{i}|X^{i}) \\
	I(X_{i+1};\widehat{\Theta}^{i}|X^{i})        &= h(X_{i+1}|X^{i}) - h(X_{i+1}|X^{i},\widehat{\Theta}^{i}) \numberthis \label{eq:dir-info-rev-noiseless}
\end{align*}
The first term in the equation above reduces to
\begin{align*}
	h(X_{i+1}|X^{i}) &= h(\Theta-\widehat{\Theta}_{i}|X^{i}) \\
					 &= h \bigg( \Theta- \bigg( \widehat{\Theta}_{i-1}+\frac{X_{i}+N_{i}}{i} \bigg) \bigg| X^{i} \bigg) \\
					 &\overset{(a)}{=} h \bigg( \Theta- \bigg( (\Theta - X_{i}) + \frac{N_{i}}{i} \bigg) \bigg| X^{i} \bigg) \\
					 &= h \bigg( \frac{N_{i}}{i} \bigg| X^{i} \bigg) \\
					 &= h(N_{i}) -\log(i) \\
					 &= \frac{1}{2}\log(2\pi e\sigma_{N}^{2}) -\log(i)
\end{align*}
where for (a), we have dropped $X_i/i$ from the previous step, since it is conditioned over, and then written $\widehat\Theta_{i-1}$ as $(\Theta - X_i)$. On the other hand, the second term in equation \eqref{eq:dir-info-rev-noiseless} becomes
\begin{align*}
	h(X_{i+1} | X^i, \widehat\Theta^i) &= h(\Theta - \widehat\Theta_i | X^i, \widehat\Theta^i) \\
									   &= h(\Theta | X^i, \widehat\Theta^i) \\
									   &\overset{(a)}{=} h(X_i + \widehat\Theta_{i-1} | X^i, \widehat\Theta^i) \\
									   &= h(0 | X^i, \widehat\Theta^i) \\
									   &\overset{(b)}{=} - \infty
\end{align*}
where for (a) we have expressed $\Theta$ in terms of $\widehat\Theta_{i-1}$ and $X_i$ and for (b), we have used the fact that the differential entropy of a constant (or equivalently, a Gaussian with zero variance) is negative infinity. This means that equation \eqref{eq:dir-info-rev-noiseless} becomes
\begin{align*}
	I(X_{i+1};\widehat{\Theta}^{i}|X^{i}) &= \infty \\
	\Rightarrow I(0*\widehat{\Theta}^{n-1}\rightarrow X^{n}) &= \infty
\end{align*}
as expected.

>>>>

\begin{document}
\title{Is the direction of greater Granger causal influence the same as the direction of information flow?}
\author{Praveen Venkatesh and Pulkit Grover\\Carnegie Mellon University}

@@

\begin{abstract}
Granger causality is an established statistical measure of the ``causal influence'' that one stochastic process has on another. Along with its more recent generalization -- Directed Information -- Granger Causality has been used extensively in neuroscience, and in complex interconnected systems in general, to infer statistical causal influences. Of late, many works have begun to interpret the direction of causal influence as the direction of ``information flow''. We ask: is the direction of greater causal influence, as predicted by Granger Causality and Directed Information always the same as the direction of information flow? We test whether comparison of these measures in either direction correctly predicts the direction of information flow by using two simple theoretical experiments, in which the true direction of information flow (the ``ground truth'') is known by design. The experiments are based on a communication system with a feedback channel, upon which a strategy inspired by the work of Schalkwijk and Kailath is employed. We show that in these experiments, the direction of information flow can be opposite to the direction of greater Granger causal influence or Directed Information. Thus, while it might be reasonable to infer the direction of statistical causal influence using these techniques, one needs to exercise care in interpreting the direction of causal influence as the direction of information flow.
\end{abstract}

@@

\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
	\centering
	\includegraphics[width=3.2in]{gc-comparison}
	\caption{In order to determine the direction of greater causal influence, the Granger Causality metrics in the forward and reverse directions are often compared~\cite{Roebroeck2005MappingDirected}.}
	\label{fig:gc-comparison}
\end{figure}

In order to determine the direction of \emph{greater} causal influence, the Granger Causality metric (ratio of residual variances) from $\{X\}$ to $\{Y\}$ is often \emph{compared} to that from $\{Y\}$ to $\{X\}$. The direction of causal influence is then taken to be the direction with the greater Granger Causality metric (e.g.~\cite{Brovelli2004BetaOscillations,jiao2013universal,Roebroeck2005MappingDirected}, see~\cite{Roebroeck2005MappingDirected} for an understanding of what physical constraints motivate this comparison)). Further, this direction of causal influence is interpreted to be the direction of information flow, which is the interpretation we question in this paper. We also note that many of these works use a spectral version of Granger Causality, that supplies this metric as a function of frequency. It then becomes possible to also determine the brain wave frequency at which these influences occur. However, we restrict our analysis to the simpler non-spectral version of Granger Causality, since it is sufficient for the purpose of our arguments.

@@

Suppose we try to use the direction of greater Granger causal influence to infer the direction of information flow. As discussed in Section~\ref{sec:intro}, one might expect the direction of information flow -- and hence the direction of greater Granger causality -- to be from the transmitter to the receiver.

@@

In the reverse direction, the Directed Information is $\infty$.
\begin{align*}
	I(0*\widehat{\Theta}^{n-1}\rightarrow X^{n}) &= \sum_{i=0}^{n-1}I(X_{i+1};\widehat{\Theta}^{i}|X^{i}) \\
	I(X_{i+1};\widehat{\Theta}^{i}|X^{i})        &= h(X_{i+1}|X^{i}) - h(X_{i+1}|X^{i},\widehat{\Theta}^{i}) \numberthis \label{eq:dir-info-rev-noiseless}
\end{align*}
The first term in the equation above reduces to
\begin{align*}
	h(X_{i+1}|X^{i}) &= h(\Theta-\widehat{\Theta}_{i}|X^{i}) \\
					 &= h \bigg( \Theta- \bigg( \widehat{\Theta}_{i-1}+\frac{X_{i}+N_{i}}{i} \bigg) \bigg| X^{i} \bigg) \\
					 &\overset{(a)}{=} h \bigg( \Theta- \bigg( (\Theta - X_{i}) + \frac{N_{i}}{i} \bigg) \bigg| X^{i} \bigg) \\
					 &= h \bigg( \frac{N_{i}}{i} \bigg| X^{i} \bigg) \\
					 &= h(N_{i}) -\log(i) \\
					 &= \frac{1}{2}\log(2\pi e\sigma_{N}^{2}) -\log(i)
\end{align*}
where for (a), we have dropped $X_i/i$ from the previous step, since it is conditioned over, and then written $\widehat\Theta_{i-1}$ as $(\Theta - X_i)$. On the other hand, the second term in equation \eqref{eq:dir-info-rev-noiseless} becomes
\begin{align*}
	h(X_{i+1} | X^i, \widehat\Theta^i) &= h(\Theta - \widehat\Theta_i | X^i, \widehat\Theta^i) \\
									   &= h(\Theta | X^i, \widehat\Theta^i) \\
									   &\overset{(a)}{=} h(X_i + \widehat\Theta_{i-1} | X^i, \widehat\Theta^i) \\
									   &= h(0 | X^i, \widehat\Theta^i) \\
									   &\overset{(b)}{=} - \infty
\end{align*}
where for (a) we have expressed $\Theta$ in terms of $\widehat\Theta_{i-1}$ and $X_i$ and for (b), we have used the fact that the differential entropy of a constant (or equivalently, a Gaussian with zero variance) is negative infinity. This means that equation \eqref{eq:dir-info-rev-noiseless} becomes
\begin{align*}
	I(X_{i+1};\widehat{\Theta}^{i}|X^{i}) &= \infty \\
	\Rightarrow I(0*\widehat{\Theta}^{n-1}\rightarrow X^{n}) &= \infty
\end{align*}
as expected.
